gratia: An R package for exploring generalized additive models

### 2024-11-21 paper.Rmd
add a state of the field section as per #304

@@

*gratia* can handle distributional GAMs fitted with *mgcv*, and also those fitted using *GJRM*'s `gamlss()`. Below, the estimated smooths from `m2` are plotted using `draw()`
```{r m2-draw, fig.show = "hide", dependson = "fit-chl-lsslmodel", cache = TRUE, fig.pos = "t!", out.width = ".8\\linewidth"}
# Define the coordinate system to use for plotting on a map
crs <- "+proj=ortho +lat_0=20 +lon_0=-40"
draw(m2, crs = crs, default_crs = 4326, dist = 0.05, rug = FALSE)
```
The plots produced by `draw()` from a fitted model are known as *partial effect plots* (Figure \ref{fig:m2-draw}), which show the component contributions, on the link scale, of each model term to the linear predictor. The y axis on these plots is typically centred around 0 due to most smooths having a sum-to-zero identifiability constraint applied to them. This constraint is what allows the model to include multiple smooths and remain identifiable. These plots allow you to read off the contributions of each smooth to the fitted response (on the link scale); they show link-scale *predictions* of the response for each smooth, conditional upon all other terms in the model, including any parametric effects and the intercept, having zero contribution. In the paralance of the *marginaleffects* package, these plots show *adjusted predictions*, just where the adjustment includes setting the contribution of *all* other model terms to the predicted value to zero. For *partial derivatives* (what *marginaleffects* would call a *marginal effect* or slope), *gratia* provides the `derivatives()`.

>>>>

# State of the field

Several R packages provide similar functionality to that of *gratia*. Notably, the *mgcViz* package [@Fasiolo2020-nz] provides sophisticated capabilities for plotting estimated smooths for models fitted by *mgcv* and *qgam* [@], and model diagnostics plots. A particular feature of *mgcViz* is that it was designed to be scalable, easily handling models fitted to data with millions of observations, a use case not currently well handled by *gratia*. *tidymv* [@tidymv-2023] and its succesor, *tidygam* [@tidygam-2023], provide plots of estimated smooths and model predictions respectively, as well as differences of smooths, but the focus is on univariate smooths, inconstrast to *gratia*'s ability to plot multivariate and specialist smooths (e.g. soap film smooths). Like *gratia*, the *itsadug* package [@itsadug-2022] is motivated to make working with estimated GAMs and related models easier. *itsadug* uses base graphics for plotting, but in place of partial effects plots, produces plots of adjusted predictions, and while useful for a broad range of models, is especially focused on GAMs fitted to longitudinal data with random effects.

Areas where *gratia* stands out from these other packages are i) the consistent functionality for range of posterior simulations (*mgcViz* has some support for simulating new response values) as illustrated below, ii) support for computing derivatives of smooths, partial derivatives, and derivatives on the response scale (i.e. conditional derivatives), iii) tools for learning or teaching how GAMs work, iv) the range of utility functions that make working with GAMs easier, and v) the simple and consistent *tidy* interface to all functionality.

@@

*gratia* can handle distributional GAMs fitted with *mgcv*, and also those fitted using *GJRM*'s `gamlss()`. Below, the estimated smooths from `m2` are plotted using `draw()`
```{r m2-draw, fig.show = "hide", dependson = "fit-chl-lsslmodel", cache = TRUE, fig.pos = "t!", out.width = ".8\\linewidth"}
# Define the coordinate system to use for plotting on a map
crs <- "+proj=ortho +lat_0=20 +lon_0=-40"
draw(m2, crs = crs, default_crs = 4326, dist = 0.05, rug = FALSE)
```
The plots produced by `draw()` from a fitted model are known as *partial effect plots* (Figure \ref{fig:m2-draw}), which show the component contributions, on the link scale, of each model term to the linear predictor. The y axis on these plots is typically centred around 0 due to most smooths having a sum-to-zero identifiability constraint applied to them. This constraint is what allows the model to include multiple smooths and remain identifiable. These plots allow you to read off the contributions of each smooth to the fitted response (on the link scale); they show link-scale *predictions* of the response for each smooth, conditional upon all other terms in the model, including any parametric effects and the intercept, having zero contribution. In the paralance of the *marginaleffects* package [@Arel-BundockGreiferHeiss:2024], these plots show *adjusted predictions*, just where the adjustment includes setting the contribution of *all* other model terms to the predicted value to zero. For *partial derivatives* (what *marginaleffects* would call a *marginal effect* or slope), *gratia* provides the `derivatives()`.
